---
title: "Reproducible Research: Peer Assessment 1"
author: "Jhon Guzman"
date: "6/23/2017"
output: html_document
keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)



getwd()
setwd("/Users/jguzman/Documents/onedrive/DataScience/gitRepo/ReproducibleResearch/Week1/Project")
list.files()


```

## Loading and preprocessing the data


### Library dependencies
```{r, include=TRUE}
library(dplyr)
library(ggplot2)
```


### Load raw data

Load raw data from `activity.csv`

```{r activityRawSet, include=TRUE}

activityRawSet <- read.csv("activity.csv",sep = ",", na.strings = "NA", stringsAsFactors = FALSE)

head(activityRawSet)

```

### Prepare data

Create a data frame from `activityRawSet`

```{r activitySet_df, include=TRUE}

activitySet_df <- tbl_df(activityRawSet)

activitySet_df <- mutate(activitySet_df, date = as.Date(activitySet_df$date, "%Y-%m-%d") )

```

Add `count_steps` variable

```{r activity_steps_taken, include=TRUE}

activity_steps_taken <- activitySet_df %>%
    group_by(date) %>%
    summarize(count_steps = sum(steps,na.rm=TRUE)
    )

activity_steps_taken

```

## What is mean total number of steps taken per day?

The mean and median of total steps taken per day were calculated, using `mean()` and `median()` functions with `activity_steps_taken$count_steps` as parameter including observations with `NA` values.

```{r , include=TRUE, echo=TRUE}

activity_mean <- mean(activity_steps_taken$count_steps)
activity_median <- median(activity_steps_taken$count_steps)

```

The mean total number of steps taken per day is:  **``r round(activity_mean, digits=2)``** and the median total number of steps taken per day is: **``r activity_median``**.

*Figure 1*, shows the histogram of the total number of steps taken each day including its total mean and median values represented by red and blue doted lines respectively.

```{r, echo=TRUE}
# Generate histogram
hist(activity_steps_taken$count_steps, breaks=25, main="Figure 1: Histogram of Total Number of Steps in each Day", xlab="Total Steps [count_steps]")
rug(activity_steps_taken$count_steps)
abline(v = activity_mean, col = "red", lwd = 1, lty = 2)
abline(v = activity_median, col = "blue", lwd = 1, lty = 2)
text(activity_mean-0.3e3, 8, bquote(bar(x) == .(activity_mean)), col = "red", srt = 90,cex=0.6)
text(activity_median+0.4e3, 8,  bquote(mu == .(activity_median)), col = "blue", srt = 90,cex=0.6)





```


## What is the average daily activity pattern?

Grouped dataset by time interval to obtain the total number and average number of steps taken per interval, averaged across all days.

```{r activity_time, include=TRUE}

# Avg number of steps grouped by interval
# create count_steps and avg_steps variables 
# include NA values
activity_time <- activitySet_df %>%
    group_by(interval) %>%
    summarize(count_steps = sum(steps,na.rm=TRUE),
              avg_steps = mean(steps,na.rm=TRUE)
    )
```

Portion of `activity_time` dataset 

```{r, include=TRUE, echo=FALSE}

activity_time

```

The summary of `activity_time$avg_steps` shows on average the maximum number of steps in the column (`Max.`) 

```{r, include=TRUE, echo=FALSE}

summary(activity_time$avg_steps)

steps_mean <- mean(activity_time$avg_steps)
steps_median <- median(activity_time$avg_steps)

```

The 5-minute interval that on average, contains the maximum number of steps is calculated as follow:

```{r max_interval, include=TRUE, echo=TRUE}

# Get the index value from the max avg_steps value of activity_time dataset 
# set result to max_interval object
max_interval<-activity_time[which.max(activity_time$avg_steps), ]
max_interval

```

Across all the days in the dataset, the 5-minute interval of **``r max_interval$interval``** minutes , with a total of **``r max_interval$count_steps``** steps, on average contains a maximum of **``r round(max_interval$avg_steps, digits=2)``** steps.

*Figure 2*, shows the time series plot of the average number of steps taken, including the 5-minute interval that on average, contains the maximum number of steps (Doted red vertical line).

```{r g2, include=TRUE, echo=TRUE}

g2 <- ggplot(activity_time, aes(interval, avg_steps))
g2 + geom_line() + ggtitle("Figure 2. Time series plot of the average number of steps taken") +  xlab("5  Interval (minutes)") +  ylab("Avg. Activity (steps)") + 
    geom_vline(aes(xintercept=max_interval$interval), color="red", linetype="dashed", size=1/2) + theme_classic()

```


## Imputing missing values

Note that there are a number of days/intervals where there are missing values (coded as ð™½ð™°). The presence of missing days may introduce bias into some calculations or summaries of the data. The total number of observations containing missing values are calculated as follow.

```{r, include=TRUE, echo=TRUE}
activity_count_na_values <- sum(is.na(activityRawSet$steps))
```

The total number of observations containing missing values in steps variable is **``r activity_count_na_values``** 

The strategy for filling in all of the missing values in the dataset is replacing NA values with the mean of 5-minute time interval and create a new dataset `activity_fill_na` that is equal to the original dataset `activitySet_df` but with the missing data filled in.

```{r activity_fill_na, include=TRUE, echo=TRUE}

# Group activitySet_df by interval
# Replace NA values in steps variable with the mean() by interval
activity_fill_na <- activitySet_df %>%
    group_by(interval)  %>%
    mutate(steps= ifelse(is.na(steps), mean(steps, na.rm=TRUE), steps))
```

Check  `activity_fill_na` dataset with 0 `NA` values

```{r , include=TRUE, echo=FALSE}
# Check NA values in dataset
colSums(is.na(activity_fill_na))
```

Portion of `activity_fill_na` dataset 

```{r , include=TRUE, echo=FALSE}
# show data 
activity_fill_na
```


## Are there differences in activity patterns between weekdays and weekends?



